---
title: "PubH7462 - Homework 1"
author: "Andrés Arguedas"
date: "2/2/2022"
output: 
  github_document:
    toc: true
---

```{r setup, include = FALSE}
# Load all the good stuff
library(tidyverse)

# Working directory for .RMD, figure output in markdown, and messages/warnings
# output
knitr::opts_knit$set(echo = TRUE,
                     root.dir = rprojroot::find_rstudio_root_file(),
                     message = FALSE,
                     warning = FALSE,
                     fig.width = 6,
                     out.width = "70%",
                     fig.align = "center",
                     cache = FALSE)

# Set Theme for ggplot2 to `theme_bw()`, as well as centering the title and
# legend at bottom by default
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5), 
             legend.position = "bottom")

# Set the color palette of ggplot to a colorblind friendly one
options(ggplot2.discrete.colour = c("#E69F00", "#56B4E9", "#009E73","#F0E442", 
                                    "#0072B2", "#D55E00", "#CC79A7", "#000000"))

# Set Scientific notation output and decimal places for knitr
options(scipen = 999)
options(digits = 4)
```

# Problem 2.1

```{r simulating-data}
sim_data <- tibble(
  x = rnorm(n = 1000, mean = 0, sd = 1),
  y = rnorm(n = 1000, mean = 1, sd = 2),
  sum_indicator = ifelse(x + y > 0.5, TRUE, FALSE)) %>% 
  mutate(indicator_TRUE = factor(sum_indicator,
                                 labels = c("No", "Yes")),
         indicator_TRUE = fct_relevel(indicator_TRUE, "Yes", "No"))
```


```{r plot-of-simulated-data}
sim_data %>% ggplot(aes(x = x, y = y, col = indicator_TRUE)) +
  geom_point() +
  labs(title = "Figure 1:
  Two random normal samples of size 1000, one with mean 0 and standard 
  deviation 1, and the other with mean 1 and standard deviation 2, 
  colored according to if their sum is greater than 0.5",
       x = expression(X %~% N(0,1)),
       y = expression(Y %~% N(1,2)),
       col = expression(X + Y > 0.5))
```

# Problem 2.2

To start, we need to load the data into R

```{r load-penguins}
# Read the `penguin.RDS` file from the `data` folder
penguin.df <- read_rds("./data/penguin.RDS")
```

## Data Description

- Observations: `r nrow(penguin.df)`
- Variables: `r ncol(penguin.df)`
- Mean flipper length: `r mean(penguin.df$flipper_length_mm, na.rm = TRUE)` mm
- Mean bill length: `r mean(penguin.df$bill_length_mm, na.rm = TRUE)` mm

| Variable            | Description                                                        | Units or levels             |
|---------------------|--------------------------------------------------------------------|-----------------------------|
| `species`           | Species of the penguin                                             | Adelie, Chinstrap or Gentoo |
| `island`            | Island of the Palmer Archipelago in which the penguin was observed | Biscoe, Dream or Torgersen  |
| `bill_length_mm`    | Penguin's bill length                                              | millimeters                 |
| `bill_depth_mm`     | Penguin's bill depth                                               | millimeters                 |
| `flipper_length_mm` | Penguin's flipper length                                           | millimeters                 |
| `body_mass_g`       | Penguin's body mass                                                | grams                       |
| `sex`               | Penguin's sex                                                      | Female or male              |
| `year`              | Study year in which the penguin was observed                       | 2007, 2008 or 2009          |

## Visualization

```{r flipper-and-bill-length-by-species, warning = FALSE}
penguin.df %>% ggplot(aes(y = flipper_length_mm, x = bill_length_mm, col = species)) +
  geom_point() +
  labs(title = "",
       x = "Bill length (mm)",
       y = "Flipper length (mm)",
       col = "Species")
```

```{r flipper-and-bill-length-by-species-and-sex, warning = FALSE}
penguin.df %>% ggplot(aes(y = flipper_length_mm, x = bill_length_mm, col = species)) +
  facet_wrap(~ sex) +
  geom_point() +
  labs(title = "",
       x = "Bill length (mm)",
       y = "Flipper length (mm)",
       col = "Species")
```